---
title: "Beta-Catenin Knockout Analysis with PathwayEmbed"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Beta-Catenin Knockout Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview
This vignette shows the application of PathwayEmbed in a beta-catenin perturbed system where Ctnnb1 upstream enhancer is depleted in instestinal epithelia. Refinements of coefficients is flexible and should be case-depedent.
References: Hua et al. A Ctnnb1 enhancer transcriptionally regulates Wnt signaling dosage to balance homeostasis and tumorigenesis of intestinal epithelia. Elife. 2024 Sep 25;13:RP98238. doi: 10.7554/eLife.98238. PMID: 39320349; PMCID: PMC11424096.

## Load Packages and Download Data from Online Source
```{r setup}
library(PathwayEmbed)
library(Seurat)

url_ko <- "https://ftp.ncbi.nlm.nih.gov/geo/series/GSE233nnn/GSE233978/suppl/GSE233978_KO_filtered_feature_bc_matrix.h5"
url_wt <- "https://ftp.ncbi.nlm.nih.gov/geo/series/GSE233nnn/GSE233978/suppl/GSE233978_WT_filtered_feature_bc_matrix.h5"


download.file(url_ko, destfile = "GSE233978_KO_filtered_feature_bc_matrix.h5", mode = "wb")
download.file(url_wt, destfile = "GSE233978_WT_filtered_feature_bc_matrix.h5", mode = "wb")
```

## Data Preparation
```{r}
# Load KO and WT expression matrices from local HDF5 files
ko_data <- Read10X_h5("GSE233978_KO_filtered_feature_bc_matrix.h5")
wt_data <- Read10X_h5("GSE233978_WT_filtered_feature_bc_matrix.h5")

# Create Seurat objects
# Apply during object creation
ko <- CreateSeuratObject(counts = ko_data, project = "KO", min.cells = 3, min.features = 200)
wt <- CreateSeuratObject(counts = wt_data, project = "WT", min.cells = 3, min.features = 200)


# Add sample metadata
ko$sample <- "KO"
wt$sample <- "WT"

# Merge and join layers
merged <- merge(ko, wt)
merged[["RNA"]] <- JoinLayers(merged[["RNA"]])
```

## Preprocessing 
Get a Seurat Object
```{r}
# Normalize and scale
merged <- NormalizeData(
  object = merged,
  normalization.method = "LogNormalize",
  scale.factor = 10000
)

merged <- FindVariableFeatures(
  object = merged,
  selection.method = "vst",
  nfeatures = 2000
)

merged <- ScaleData(
  object = merged,
  features = VariableFeatures(object = merged)
)

```

## Wnt Pathway Scoring Using Pathway Embed
```{r}
# Load Pathway database (can select one or use all)
Wnt_12h <- LoadPathway("WNT3A_12H_ACTIVATION", "mouse")
Wnt_24h <- LoadPathway("WNT3A_24H_ACTIVATION", "mouse")
Wnt_48h <- LoadPathway("WNT3A_48H_ACTIVATION", "mouse")
Wnt_slope <- LoadPathway("WNT3A_SLOPE_ACTIVATION","mouse")

# Data Preprocess to get a standardized matrix (z-scored matrix for pathway genes x cells)
matrix_12h <- DataPreProcess(merged, Wnt_12h, Seurat.object = TRUE)
matrix_24h<-DataPreProcess(merged, Wnt_24h, Seurat.object = TRUE)
matrix_48h<-DataPreProcess(merged, Wnt_48h, Seurat.object = TRUE)
matrix_slope<-DataPreProcess(merged, Wnt_slope, Seurat.object = TRUE)

# Pathway Max and Min for this dataset
pathwaystat_12h <- PathwayMaxMin(matrix_12h, Wnt_12h)
pathwaystat_24h <- PathwayMaxMin(matrix_24h, Wnt_24h)
pathwaystat_48h <- PathwayMaxMin(matrix_12h, Wnt_48h)
pathwaystat_slope <- PathwayMaxMin(matrix_12h, Wnt_slope)

# Calculate the score for each cell
score_12h <- ComputeCellData(matrix_12h, pathwaystat_12h)
score_24h <- ComputeCellData(matrix_24h, pathwaystat_24h)
score_48h <- ComputeCellData(matrix_48h, pathwaystat_48h)
score_slope <- ComputeCellData(matrix_slope, pathwaystat_slope)
```

## Visualization 
Use functions included in the packages first
```{r}
# Format score results and metadata for plotting
plot_data_12h <- PreparePlotData(merged, score_12h, "sample", Seurat.object = TRUE)
plot_data_24h <- PreparePlotData(merged, score_24h, "sample", Seurat.object = TRUE)
plot_data_48h <- PreparePlotData(merged, score_48h, "sample", Seurat.object = TRUE)
plot_data_slope <- PreparePlotData(merged, score_slope, "sample", Seurat.object = TRUE)

# Check data
CalculatePercentage(to.plot = plot_data_12h,group_var = "sample")
CalculatePercentage(to.plot = plot_data_24h,group_var = "sample")
CalculatePercentage(to.plot = plot_data_48h,group_var = "sample")
CalculatePercentage(to.plot = plot_data_slope,group_var = "sample")

# Use slope data to visualize 
# Extract Effect Size
pct_slope <- CalculatePercentage(to.plot = plot_data_slope, group_var = "sample")
cohens_d_slope <- round(pct_slope$cohens_d[1], 3)

# Density Plot using the default function and annotation

PlotPathway(
  to.plot = plot_data_slope,
  pathway = "Wnt",
  group = "sample",
  color = c("#ae282c", "#2066a8")
) +
  annotate(
    "text",
    x     = Inf, y = Inf,
    label = paste0("Cohen's d = ", cohens_d_slope),
    hjust = 1.1, vjust = 1.5,
    size  = 3.5,
    color = "black"
  ) +
  xlim(-5, 5)

```
