---
title: "Toy Set"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Toy Set}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview
This vignette demonstrates how to use the PathwayEmbed package to compute and visualize pathway activation using single-cell transcriptomic data. We use the example dataset synthetic_test_object included with the package.

## Load Package and Example Data
```{r setup}
library(PathwayEmbed)
library(dplyr)
library(tidyr)
library(ggridges)
# Load the example gene expression matrix included in the package
data(synthetic_test_matrix)
```

## Check Pathway Data
```{r}
# List all information in the Pathway database
ListPathway()
# List all available pathways
ListPathway("Pathway")
# List all options related to "WNT" pathway
ListPathway("WNT")
# Load Pathway database (can select one or use all)
Wnt_12h <- LoadPathway("WNT3A_12H_ACTIVATION")
Wnt_24h <- LoadPathway("WNT3A_24H_ACTIVATION")
Wnt_48h <- LoadPathway("WNT3A_48H_ACTIVATION")
Wnt_slope <- LoadPathway("WNT3A_SLOPE_ACTIVATION")
```

## Check Pathway Data
```{r}
# List all information in the Pathway database
ListPathway()
# List all available pathways
ListPathway("Pathway")
# List all options related to "WNT" pathway
ListPathway("WNT")
# Load Pathway database (can select one or use all)
Wnt_12h <- LoadPathway("WNT3A_12H_ACTIVATION", "mouse")
Wnt_24h <- LoadPathway("WNT3A_24H_ACTIVATION", "mouse")
Wnt_48h <- LoadPathway("WNT3A_48H_ACTIVATION", "mouse")
Wnt_slope <- LoadPathway("WNT3A_SLOPE_ACTIVATION","mouse")
```

## Data Preprocess and Evaluation (pathway max and min)
```{r}

# Data Preprocess to get a standardized matrix (z-scored matrix for pathway genes x cells)
matrix_12h <- DataPreProcess(synthetic_test_matrix, Wnt_12h, Seurat.object = FALSE)
matrix_24h<-DataPreProcess(synthetic_test_matrix, Wnt_24h, Seurat.object = FALSE)
matrix_48h<-DataPreProcess(synthetic_test_matrix, Wnt_48h, Seurat.object = FALSE)
matrix_slope<-DataPreProcess(synthetic_test_matrix, Wnt_slope, Seurat.object = FALSE)

# Pathway Max and Min for this dataset
pathwaystat_12h <- PathwayMaxMin(matrix_12h, Wnt_12h)
pathwaystat_24h <- PathwayMaxMin(matrix_24h, Wnt_24h)
pathwaystat_48h <- PathwayMaxMin(matrix_12h, Wnt_48h)
pathwaystat_slope <- PathwayMaxMin(matrix_12h, Wnt_slope)

```

## Calculate Pathway Score for each cell
```{r}
# Calculate the score for each cell
score_12h <- ComputeCellData(matrix_12h, pathwaystat_12h)
score_24h <- ComputeCellData(matrix_24h, pathwaystat_24h)
score_48h <- ComputeCellData(matrix_48h, pathwaystat_48h)
score_slope <- ComputeCellData(matrix_slope, pathwaystat_slope)
```

## Prepare Data for Plotting
```{r}
# Get metadata
data("synthetic_test_metadata")
# Format score results and metadata for plotting
plot_data_12h <- PreparePlotData(synthetic_test_metadata,score_12h,group = "genotype")
plot_data_24h <- PreparePlotData(synthetic_test_metadata,score_24h,group = "genotype")
plot_data_48h <- PreparePlotData(synthetic_test_metadata,score_48h,group = "genotype")
plot_data_slope <- PreparePlotData(synthetic_test_metadata,score_slope,group = "genotype")
```

## Visualize Pathway Activation Using the Function Included
The classic pathway output from `PlotPathway()`. Each curve represents the distribution of z-scored WNT activation across all cells in a group. The dotted vertical line at 0 divides ON (right) from OFF (left) cells.
```{r}
# Visualize pathway activation colored by genotype
PlotPathway(
  to.plot = plot_data_12h,
  pathway = "12hr Wnt",
  group = "genotype",
  color = c("#ae282c", "#2066a8")
)
PlotPathway(
  to.plot = plot_data_24h,
  pathway = "24hr Wnt",
  group = "genotype",
  color = c("#ae282c", "#2066a8")
)
PlotPathway(
  to.plot = plot_data_48h,
  pathway = "48hr Wnt",
  group = "genotype",
  color = c("#ae282c", "#2066a8")
)
PlotPathway(
  to.plot = plot_data_slope,
  pathway = "Wnt",
  group = "genotype",
  color = c("#ae282c", "#2066a8")
)
```

## Calculate Group-wise Activation Percentage (Optional)
```{r}
# Calculate % of cells per group with high pathway activation
CalculatePercentage(to.plot = plot_data_12h,group_var = "genotype")
CalculatePercentage(to.plot = plot_data_24h,group_var = "genotype")
CalculatePercentage(to.plot = plot_data_48h,group_var = "genotype")
CalculatePercentage(to.plot = plot_data_slope,group_var = "genotype")
```
## Add Cohen's d As a Text Annotation on the Density Plot
```{r}
# Extract Cohen's d from one of the timepoints
pct_12h <- CalculatePercentage(to.plot = plot_data_12h, group_var = "genotype")
cohens_d_12h <- round(pct_12h$cohens_d[1], 3)

# Add it as a label on the plot
PlotPathway(
  to.plot = plot_data_12h,
  pathway = "12hr Wnt",
  group = "genotype",
  color = c("#ae282c", "#2066a8")
)+
  annotate(
    "text",
    x     = Inf, y = Inf,
    label = paste0("Cohen's d = ", cohens_d_12h),
    hjust = 1.1, vjust = 1.5,
    size  = 3.5,
    color = "black"
  )
```

## Other Visualization 2 — Violin + Jitter Plot

Violins convey the full distribution shape while also reflecting relative cell abundance through width. Overlaid jittered points expose individual cell-level heterogeneity — particularly useful when group sizes differ substantially.

```{r }
ggplot(plot_data_12h, aes(x = genotype, y = scale, fill = genotype, color = genotype)) +
  geom_violin(alpha = 0.5, trim = FALSE) +
  geom_jitter(width = 0.15, size = 0.6, alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "black", linewidth = 0.5) +
  scale_fill_manual(values  =  c("#ae282c", "#2066a8")) +
  scale_color_manual(values =  c("#ae282c", "#2066a8")) +
  labs(
    title = "WNT Pathway Transduction",
    x     = NULL,
    y     = "Relative Transduction State (z-score)"
  ) +
  theme_classic() +
  theme(legend.position = "none")
```

---

## Other Visualization 3 — Ridgeline Plot Across Timepoints

Stacked ridgeline plots (`ggridges`) reveal how the activation distribution evolves across the four WNT time points. Each ridge is a separate density; rows are ordered chronologically so temporal trends read top-to-bottom.

```{r}
# Combine all time points into one data frame
ridge_data <- bind_rows(
  mutate(plot_data_12h,   timepoint = "12h"),
  mutate(plot_data_24h,   timepoint = "24h"),
  mutate(plot_data_48h,   timepoint = "48h"),
  mutate(plot_data_slope, timepoint = "Slope")
) %>%
  mutate(timepoint = factor(timepoint, levels = c("12h", "24h", "48h", "Slope")))
group_colors <- c("WT" = "#ae282c", "KO" = "#2066a8")
ggplot(ridge_data, aes(x = scale, y = timepoint, fill = genotype)) +
  geom_density_ridges(alpha = 0.55, scale = 1.2, rel_min_height = 0.01) +
  geom_vline(xintercept = 0, linetype = "dotted", color = "black", linewidth = 0.5) +
  scale_fill_manual(values = group_colors) +
  labs(
    title = "WNT3A Activation Across Timepoints",
    x     = "Relative Transduction State (z-score)",
    y     = "Timepoint"
  ) +
  theme_classic()
```

---

## Visualization 4 — ON/OFF Stacked Bar Chart

`CalculatePercentage()` quantifies how many cells cross the z = 0 threshold. A stacked bar chart makes the ON/OFF balance immediately legible — especially useful for comparing multiple groups or timepoints at a glance.

```{r viz4-bar, fig.cap = "Percentage of ON vs OFF cells per group at each WNT timepoint."}
# Calculate ON/OFF percentages for each time point
pct_list <- list(
  "12h"   = CalculatePercentage(plot_data_12h,   "genotype"),
  "24h"   = CalculatePercentage(plot_data_24h,   "genotype"),
  "48h"   = CalculatePercentage(plot_data_48h,   "genotype"),
  "Slope" = CalculatePercentage(plot_data_slope, "genotype")
)

pct_df <- bind_rows(pct_list, .id = "timepoint") %>%
  pivot_longer(cols = c(percentage_on, percentage_off),
               names_to  = "state",
               values_to = "percentage") %>%
  mutate(
    timepoint = factor(timepoint, levels = c("12h", "24h", "48h", "Slope")),
    state     = recode(state, percentage_on = "ON", percentage_off = "OFF"),
    state     = factor(state, levels = c("ON", "OFF"))
  )

ggplot(pct_df, aes(x = timepoint, y = percentage, fill = state)) +
  geom_bar(stat = "identity", position = "stack", color = "white", linewidth = 0.3) +
  facet_wrap(~ group) +
  scale_fill_manual(values = c("ON" = "#ae282c", "OFF" = "#2066a8")) +
  labs(
    title = "WNT Pathway ON/OFF State Compared to WNT3A Activation",
    x     = "Different activation timepoints",
    y     = "Percentage of Cells (%)",
    fill  = "State"
  ) +
  theme_classic() +
  theme(strip.background = element_rect(fill = "grey90", color = NA))
```

---

## Visualization 5 — Dot Plot: Effect Size vs. Mean Activation Shift

When comparing two groups, Cohen's d from `CalculatePercentage()` quantifies the effect size. Plotting Cohen's d (x-axis) against the mean z-score difference (y-axis) across timepoints lets you distinguish large-effect timepoints from subtle ones, with dot size encoding total cell count.

```{r viz5-dotplot, fig.cap = "Effect size (Cohen's d) vs. mean activation difference across WNT timepoints."}
# Collect Cohen's d and mean z-score difference per timepoint
effect_df <- bind_rows(
  lapply(list("12h" = plot_data_12h, "24h" = plot_data_24h,
              "48h" = plot_data_48h, "Slope" = plot_data_slope),
    function(df) {
      groups  <- unique(df$genotype)
      g1_vals <- df$scale[df$genotype == groups[1]]
      g2_vals <- df$scale[df$genotype == groups[2]]
      data.frame(
        cohens_d   = CalculatePercentage(df, "genotype")$cohens_d[1],
        mean_diff  = mean(g1_vals, na.rm = TRUE) - mean(g2_vals, na.rm = TRUE),
        total_cells = nrow(df)
      )
    }),
  .id = "timepoint"
) %>%
  mutate(timepoint = factor(timepoint, levels = c("12h", "24h", "48h", "Slope")))

ggplot(effect_df, aes(x = cohens_d, y = mean_diff,
                       size = total_cells, color = timepoint, label = timepoint)) +
  geom_point(alpha = 0.8) +
  geom_text(vjust = -1.2, size = 3.5, show.legend = FALSE) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey60") +
  scale_size_continuous(range = c(4, 10), name = "Total Cells") +
  labs(
    title  = "Effect Size vs. Mean Activation Shift — WNT Timepoints",
    x      = "Cohen's d",
    y      = "Mean z-score Difference (Group 1 − Group 2)",
    color  = "Timepoint"
  ) +
  theme_classic()
```

---

## Visualization 6 — Waterfall Plot (Ranked Cell Activation)

For a single group, cells are ranked by their z-score and plotted as a bar chart. Bars above zero are colored red (ON), bars below are blue (OFF). This reveals inter-cell heterogeneity within a condition and is especially effective for highlighting bimodal populations.

```{r}
# Use the 48h timepoint; rank cells within each group
waterfall_data <- plot_data_48h %>%
  group_by(genotype) %>%
  arrange(scale, .by_group = TRUE) %>%
  mutate(
    rank  = row_number(),
    state = ifelse(scale >= 0, "ON", "OFF")
  ) %>%
  ungroup()

ggplot(waterfall_data, aes(x = rank, y = scale, fill = state)) +
  geom_bar(stat = "identity", width = 1) +
  facet_wrap(~ genotype, scales = "free_x") +
  scale_fill_manual(values = c("ON" = "#ae282c", "OFF" = "#2066a8")) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.4) +
  labs(
    title = "Waterfall Plot of Cell Activation",
    x     = "Cells (ranked by activation)",
    y     = "Relative Transduction State (z-score)",
    fill  = "State"
  ) +
  theme_classic() +
  theme(
    axis.text.x  = element_blank(),
    axis.ticks.x = element_blank(),
    strip.background = element_rect(fill = "grey90", color = NA)
  )
```

---
---

