---
title: "Toy Set"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Toy Set}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview
This vignette demonstrates how to use the PathwayEmbed package to compute and visualize pathway activation using single-cell transcriptomic data. We use the example dataset synthetic_test_object included with the package.

## Load Package and Example Data
```{r setup}
library(PathwayEmbed)
library(dplyr)
library(tidyr)
library(ggridges)

# Load the example gene expression matrix included in the package
data(synthetic_test_matrix)
```

## Check Pathway Data
```{r}
# List all information in the Pathway database
ListPathway()
# List all available pathways
ListPathway("Pathway")
# List all options related to "WNT" pathway
ListPathway("WNT")
# Load Pathway database (can select one or use all)
Wnt_12h <- LoadPathway("WNT3A_12H_ACTIVATION", "mouse")
Wnt_24h <- LoadPathway("WNT3A_24H_ACTIVATION", "mouse")
Wnt_48h <- LoadPathway("WNT3A_48H_ACTIVATION", "mouse")
Wnt_slope <- LoadPathway("WNT3A_SLOPE_ACTIVATION","mouse")
```

## Data Preprocess and Evaluation (pathway max and min)
```{r, echo=FALSE}
synthetic_test_matrix <- fake_test_matrix
```

```{r}

# Data Preprocess to get a standardized matrix (z-scored matrix for pathway genes x cells)
matrix_12h <- DataPreProcess(synthetic_test_matrix, Wnt_12h, Seurat.object = FALSE)
matrix_24h<-DataPreProcess(synthetic_test_matrix, Wnt_24h, Seurat.object = FALSE)
matrix_48h<-DataPreProcess(synthetic_test_matrix, Wnt_48h, Seurat.object = FALSE)
matrix_slope<-DataPreProcess(synthetic_test_matrix, Wnt_slope, Seurat.object = FALSE)

# Pathway Max and Min for this dataset
pathwaystat_12h <- PathwayMaxMin(matrix_12h, Wnt_12h)
pathwaystat_24h <- PathwayMaxMin(matrix_24h, Wnt_24h)
pathwaystat_48h <- PathwayMaxMin(matrix_48h, Wnt_48h)
pathwaystat_slope <- PathwayMaxMin(matrix_slope, Wnt_slope)

```

## Calculate Pathway Score for each cell
```{r}
# Calculate the score for each cell
score_12h <- ComputeCellData(matrix_12h, pathwaystat_12h, percentile = 1)
score_24h <- ComputeCellData(matrix_24h, pathwaystat_24h, percentile = 1)
score_48h <- ComputeCellData(matrix_48h, pathwaystat_48h, percentile = 1)
score_slope <- ComputeCellData(matrix_slope, pathwaystat_slope, percentile = 1)
```

## Prepare Data for Plotting
```{r}
# Get metadata
data("synthetic_test_metadata")

# Format score results and metadata for plotting
plot_data_12h <- PreparePlotData(synthetic_test_metadata,score_12h,group = "genotype")
plot_data_24h <- PreparePlotData(synthetic_test_metadata,score_24h,group = "genotype")
plot_data_48h <- PreparePlotData(synthetic_test_metadata,score_48h,group = "genotype")
plot_data_slope <- PreparePlotData(synthetic_test_metadata,score_slope,group = "genotype")
```

## Visualize Pathway Activation Using the Function Included
The classic pathway output from `PlotPathway()`. Each curve represents the distribution of z-scored WNT activation across all cells in a group. The dotted vertical line at 0 divides ON (right) from OFF (left) cells.
```{r}
# Visualize pathway activation colored by genotype
PlotPathway(
  to.plot = plot_data_12h,
  pathway = "12hr Wnt",
  group = "genotype",
  color = c("#ae282c", "#2066a8")
)
PlotPathway(
  to.plot = plot_data_24h,
  pathway = "24hr Wnt",
  group = "genotype",
  color = c("#ae282c", "#2066a8")
)
PlotPathway(
  to.plot = plot_data_48h,
  pathway = "48hr Wnt",
  group = "genotype",
  color = c("#ae282c", "#2066a8")
)
PlotPathway(
  to.plot = plot_data_slope,
  pathway = "Wnt",
  group = "genotype",
  color = c("#ae282c", "#2066a8")
)
```

## Calculate Group-wise Activation Percentage (Optional)
```{r}
# Calculate % of cells per group with high pathway activation
CalculatePercentage(to.plot = plot_data_12h,group_var = "genotype")
CalculatePercentage(to.plot = plot_data_24h,group_var = "genotype")
CalculatePercentage(to.plot = plot_data_48h,group_var = "genotype")
CalculatePercentage(to.plot = plot_data_slope,group_var = "genotype")
```
## Add Cohen's d As a Text Annotation on the Density Plot
```{r}
# Extract Cohen's d from one of the timepoints
pct_12h <- CalculatePercentage(to.plot = plot_data_12h, group_var = "genotype")
cohens_d_12h <- round(pct_12h$cohens_d[1], 3)
p_val_12h    <- formatC(pct_12h$p_value[1], format = "e", digits = 2)

# Build per-group percentage labels
wt_label  <- paste0("WT: ON=",     pct_12h$percentage_on[pct_12h$group == "WT"],     "%, OFF=", pct_12h$percentage_off[pct_12h$group == "WT"],     "%")
mut_label <- paste0("Mutant: ON=", pct_12h$percentage_on[pct_12h$group == "Mutant"], "%, OFF=", pct_12h$percentage_off[pct_12h$group == "Mutant"], "%")

PlotPathway(
  to.plot = plot_data_12h,
  pathway = "12hr Wnt",
  group   = "genotype",
  color   = c("#ae282c", "#2066a8")
) +
  annotate("text",
    x = Inf, y = Inf,
    label = paste0("Cohen's d = ", cohens_d_12h, "\np = ", p_val_12h, "\n", wt_label, "\n", mut_label),
    hjust = 1.1, vjust = 1.5,
    size  = 3.5,
    color = "black"
  )
```

This will produce a single annotation block like:
```
Cohen's d = -0.223
p = 5.18e-09
WT: ON=45%, OFF=55%
Mutant: ON=57.5%, OFF=42.5%
```

## Other Visualization 2 — Violin + Jitter Plot

Violins convey the full distribution shape while also reflecting relative cell abundance through width. Overlaid jittered points expose individual cell-level heterogeneity — particularly useful when group sizes differ substantially.

```{r }
ggplot(plot_data_12h, aes(x = genotype, y = scale, fill = genotype, color = genotype)) +
  geom_violin(alpha = 0.5, trim = FALSE) +
  geom_jitter(width = 0.15, size = 0.6, alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "black", linewidth = 0.5) +
  scale_fill_manual(values  = c("#ae282c", "#2066a8")) +
  scale_color_manual(values = c("#ae282c", "#2066a8")) +
  labs(
    title = "WNT Pathway Transduction",
    x     = NULL,
    y     = "Relative Transduction State (z-score)"
  ) +
  theme_classic() +
  theme(legend.position = "none") +
  annotate("text",
    x     = Inf, y = Inf,
    label = paste0("Cohen's d = ", cohens_d_12h, "\np = ", p_val_12h, "\n", wt_label, "\n", mut_label),
    hjust = 1.1, vjust = 1.5,
    size  = 3.5,
    color = "black"
  )
```


---

## Visualization 3 — Waterfall Plot (Ranked Cell Activation)

For a single group, cells are ranked by their z-score and plotted as a bar chart. Bars above zero are colored red (ON), bars below are blue (OFF). This reveals inter-cell heterogeneity within a condition and is especially effective for highlighting bimodal populations.

```{r}
# rank cells within each group
waterfall_data <- plot_data_12h %>%
   group_by(genotype) %>%
   arrange(scale, .by_group = TRUE) %>%
   mutate(
     rank  = row_number(),
     state = ifelse(scale >= 0, "ON", "OFF")
   ) %>%   
  ungroup()
# Build annotation data frame with one row per genotype
annotation_df <- pct_12h %>%
  mutate(
    label = paste0(
      "Cohen's d = ", round(cohens_d, 3), "\n",
      "p = ",         formatC(p_value, format = "e", digits = 2), "\n",
      "ON = ",        percentage_on,  "%\n",
      "OFF = ",       percentage_off, "%"
    ),
    genotype = group  # must match facet variable name
  )

ggplot(waterfall_data, aes(x = rank, y = scale, fill = state)) +
  geom_bar(stat = "identity", width = 1) +
  facet_wrap(~ genotype, scales = "free_x") +
  scale_fill_manual(values = c("ON" = "#ae282c", "OFF" = "#2066a8")) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.4) +
  geom_text(
    data    = annotation_df,
    aes(x = Inf, y = Inf, label = label),
    hjust   = 1.1, vjust = 1.5,
    size    = 3,
    color   = "black",
    inherit.aes = FALSE
  ) +
  labs(
    title = "Waterfall Plot of Cell Activation",
    x     = "Cells (ranked by activation)",
    y     = "Relative Transduction State (z-score)",
    fill  = "State"
  ) +
  theme_classic() +
  theme(
    axis.text.x      = element_blank(),
    axis.ticks.x     = element_blank(),
    strip.background = element_rect(fill = "grey90", color = NA)
  )
```


## Different Percentile of Pathway on and off
When use different percentile of reference mat, would this resolve some issue of extreme on and off?
```{r }
percents <- c(1, 0.9, 0.8, 0.7, 0.6, 0.5, 0.4, 0.3, 0.2, 0.1)
outcome_list <- list()
cohens_d_percents <- list()

for (i in seq_along(percents)){
  percent <- percents[[i]]
  score <- ComputeCellData(matrix_24h, pathwaystat_24h, percentile = percent)
  plot_data <- PreparePlotData(synthetic_test_metadata, score,group = "genotype")
  outcome <- CalculatePercentage(to.plot = plot_data, group_var = "genotype")
  outcome_list[[i]] <- outcome
  cohens_d_percents[[i]] <- outcome$cohens_d[1]
}

df <- data_frame(percents = percents, 
                 cohens_d = unlist(cohens_d_percents))
print(df)

library(ggplot2)

ggplot(df, aes(x = percents, y = abs(cohens_d))) +
  geom_line(color = "#534AB7", linewidth = 0.8) +
  geom_point(color = "#534AB7", size = 2.5, fill = "white", shape = 21, stroke = 1.5) +
  geom_hline(yintercept = 0.8, linetype = "dashed", color = "#E24B4A", linewidth = 0.6) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "#EF9F27", linewidth = 0.6) +
  annotate("text", x = max(df$percents), y = 0.82, label = "large (0.8)",
           hjust = 1, size = 3, color = "#E24B4A") +
  annotate("text", x = max(df$percents), y = 0.52, label = "medium (0.5)",
           hjust = 1, size = 3, color = "#EF9F27") +
  scale_x_continuous(labels = scales::percent) +
  labs(x = "Percentile", y = "|Cohen's d|") +
  theme_minimal()

```



## Null Distribution with different gene numbers
Do random genes of the same set size output similar values?
```{r }
set.seed(123)

all_genes <- rownames(matrix_24h) # actually use the genes in the datasets to avoid no gene overlap scenario
n_genes <- length(all_genes) #15
n_reps <- 3                  # repetitions per gene set size

outcome_gene_list  <- list()
cohens_d_genes     <- data.frame()

for (i in 2:n_genes) {
  for (rep in seq_len(n_reps)) {

    # Randomly sample i genes each repetition
    sampled_genes <- sample(all_genes, i)
    expr_matrix  <- matrix_24h[sampled_genes, ]
    pathwaystat  <- PathwayMaxMin(expr_matrix, Wnt_24h)
    score        <- ComputeCellData(expr_matrix, pathwaystat, percentile = 0.5) # choose 0.5 as this shows to contribute to largest effect size
    plot_data    <- PreparePlotData(synthetic_test_metadata, score,
                                   group = "genotype")
    outcome_gene <- CalculatePercentage(to.plot = plot_data,
                                        group_var = "genotype")

    cohens_d_genes <- rbind(cohens_d_genes, data.frame(
      n_genes  = i,
      rep      = rep,
      cohens_d = outcome_gene$cohens_d[1]
    ))
  }
}
print(cohens_d_genes)

# Visualize
library(ggplot2)
library(dplyr)

cohens_d_summary <- cohens_d_genes |>
  group_by(n_genes) |>
  summarise(
    mean_d = mean(cohens_d, na.rm = TRUE),
    sd_d   = sd(cohens_d,   na.rm = TRUE),
    .groups = "drop"
  )

ggplot() +
  geom_hline(yintercept = 0, linetype = "solid",
             color = "#888780", linewidth = 0.4) +
  geom_point(data = cohens_d_genes,
             aes(x = n_genes, y = cohens_d),
             color = "#534AB7", alpha = 0.25, size = 1.2) +
  geom_ribbon(data = cohens_d_summary,
              aes(x = n_genes, ymin = mean_d - sd_d, ymax = mean_d + sd_d),
              fill = "#534AB7", alpha = 0.15) +
  geom_line(data = cohens_d_summary,
            aes(x = n_genes, y = mean_d),
            color = "#534AB7", linewidth = 0.9) +
  geom_hline(yintercept =  0.8, linetype = "dashed",
             color = "#E24B4A", linewidth = 0.5) +
  geom_hline(yintercept = -0.8, linetype = "dashed",
             color = "#E24B4A", linewidth = 0.5) +
  geom_hline(yintercept =  0.5, linetype = "dashed",
             color = "#EF9F27", linewidth = 0.5) +
  geom_hline(yintercept = -0.5, linetype = "dashed",
             color = "#EF9F27", linewidth = 0.5) +
  annotate("text", x = 15, y =  0.82, label = "large (+0.8)",  hjust = 1, size = 3, color = "#E24B4A") +
  annotate("text", x = 15, y = -0.82, label = "large (−0.8)",  hjust = 1, size = 3, color = "#E24B4A") +
  labs(x = "Number of genes", y = "Cohen's d") +
  theme_minimal()
```
## Correlation with technical cofounders (Mean/z-score)
if simply take the mean or Z-scored value of genes, is this highly correlated with the ComputeCellData output?
○ How correlated are output values to these technical confounders?
○ How similar are output values between pathways (e.g., via Spearman correlation)? If pathways share a high Spearman, this could be indicative of confounding.
```{r }
# --- Mean and Z-score summaries per cell ---
common_genes <- intersect(rownames(synthetic_test_matrix), Wnt_24h$Gene_Symbol)
gene_matrix  <- synthetic_test_matrix[common_genes, ]

mean_expr    <- colMeans(gene_matrix, na.rm = TRUE)
gene_zscored <- t(scale(t(gene_matrix)))
zscore_mean_per_cell <- colMeans(gene_zscored, na.rm = TRUE)

# --- ComputeCellData score for Wnt and get scale and cohens'd ---
wnt_score    <- ComputeCellData(matrix_24h, pathwaystat_24h, percentile = 0.6)
plot_data <- PreparePlotData(synthetic_test_metadata, wnt_score, group = "genotype")
outcome <- CalculatePercentage(to.plot = plot_data, group_var = "genotype")

# --- Combine and correlate ---
confounder_df <- data.frame(
  cell        = names(wnt_score),
  compute_score = as.numeric(wnt_score),
  normalized_scale = as.numeric(plot_data$scale),
  mean_expr   = mean_expr[names(wnt_score)],
  zscore_expr = zscore_mean_per_cell[names(wnt_score)]
)

cor_mean   <- cor(confounder_df$compute_score, confounder_df$mean_expr,
                  method = "spearman")
cor_zscore <- cor(confounder_df$compute_score, confounder_df$zscore_expr,
                  method = "spearman")

cat("Spearman correlation with mean expression:", round(cor_mean,   3), "\n")
cat("Spearman correlation with Z-score mean:   ", round(cor_zscore, 3), "\n")

# --- Scatter plots ---
library(ggplot2)
library(patchwork)

p1 <- ggplot(confounder_df, aes(x = mean_expr, y = compute_score)) +
  geom_point(alpha = 0.4, size = 1.2, color = "#534AB7") +
  geom_smooth(method = "lm", color = "#E24B4A", linewidth = 0.8, se = TRUE) +
  labs(title = sprintf("Mean expression  (ρ = %.3f)", cor_mean),
       x = "Mean gene expression", y = "ComputeCellData score") +
  theme_minimal()

p2 <- ggplot(confounder_df, aes(x = zscore_expr, y = compute_score)) +
  geom_point(alpha = 0.4, size = 1.2, color = "#534AB7") +
  geom_smooth(method = "lm", color = "#E24B4A", linewidth = 0.8, se = TRUE) +
  labs(title = sprintf("Z-score mean  (ρ = %.3f)", cor_zscore),
       x = "Mean Z-score", y = "ComputeCellData score") +
  theme_minimal()

p1 + p2

```
## Benchmark
AddModuleScore will be a good one but not apply to the example datasets as this one is small and generated specific for Wnt mutation. Therefore, benckmark to Progeny
```{r Benchmark}
library(progeny)

progeny_scores <- progeny(
  synthetic_test_matrix,
  scale    = FALSE,
  organism = "Mouse",
  top      = 100,
  perm     = 1
)

progeny_wnt <- progeny_scores[, "WNT"]  # column name depends on PROGENy pathway names

# Compare all methods
benchmark_df <- data.frame(
  cell        = names(wnt_score),
  compute_score = as.numeric(wnt_score),
  progeny_score = progeny_wnt[names(wnt_score)],
  mean_expr     = mean_expr[names(wnt_score)],
  zscore_expr = zscore_mean_per_cell[names(wnt_score)]
)

# Pairwise Spearman across all methods 
method_cols <- c("compute_score", "progeny_score",
                 "mean_expr", "zscore_expr")
benchmark_cor <- cor(benchmark_df[, method_cols], method = "spearman",
                     use = "pairwise.complete.obs")
print(benchmark_cor)

library(pheatmap)
pheatmap(
  benchmark_cor,
  color            = colorRampPalette(c("#185FA5", "white", "#993C1D"))(100),
  breaks           = seq(-1, 1, length.out = 101),
  display_numbers  = TRUE,
  number_format    = "%.2f",
  fontsize_number  = 9,
  main             = "Method benchmark — Spearman correlation"
)
```


---
---

