---
title: "Notch_Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Notch_Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
## Overview
This vignette demonstrates the application of PathwayEmbed in a Notch pertubed system (Ncstnfl/fl (cKO) bone marrow) and aging SSPC. The data reference: Remark, L.H., Leclerc, Ramsukh et al. Loss of Notch signaling in skeletal stem cells enhances bone formation with aging. Bone Res 11, 50 (2023). https://doi.org/10.1038/s41413-023-00283-8

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Load Packages and Download Data from Online Source

```{r setup}
library(Seurat)
library(PathwayEmbed)
```

## Data Preprocess For WT and KO cells

```{r, eval=FALSE}
# Load WT and KO downloaded
wt_data <- Read10X(data.dir = "WT", gene.column = 2)
ko_data <- Read10X(data.dir = "KO", gene.column = 2)

# Create Seurat objects
wt_seurat <- CreateSeuratObject(counts = wt_data, project = "WT")
ko_seurat <- CreateSeuratObject(counts = ko_data, project = "KO")

# Add metadata
wt_seurat$condition <- "WT"
ko_seurat$condition <- "KO"

# Optional: merge
combined <- merge(wt_seurat, y = ko_seurat, add.cell.ids = c("WT", "KO"), project = "WT_vs_KO")
combined[['RNA']] <- JoinLayers(combined[['RNA']])

# QC
# Add % mitochondrial content
combined[["percent.mt"]] <- PercentageFeatureSet(combined, pattern = "^mt-")
VlnPlot(combined, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
combined <- subset(combined, subset = nFeature_RNA > 200 & nFeature_RNA < 2000 & percent.mt < 10)
# Before filtering (already run above, so you can skip if done)
VlnPlot(combined, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"))

# Count cells per condition after filtering
table(combined$condition)
# KO   WT
# 8690 7778

# Normalization and Scale
combined <- NormalizeData(combined, normalization.method = "LogNormalize", scale.factor = 10000)
combined <- FindVariableFeatures(combined, selection.method = "vst", nfeatures = 2000)
combined <- ScaleData(combined)

# PCA
combined <- RunPCA(combined)
ElbowPlot(combined,ndims = 50)  # Check how many PCs to use
combined <- FindNeighbors(combined, dims = 1:20)
combined <- FindClusters(combined, resolution = 1)
combined <- RunUMAP(combined, dims = 1:20)

# Visualize
DimPlot(combined, label = TRUE)
FeaturePlot(combined, "Cxcl12")
FeaturePlot(combined, "Kitl")# 23, 25 SSPC Marker in the paper

COI_2 <- subset(combined, subset = seurat_clusters %in% c("23", "25")) # SSPC
```

## Notch Signaling Examination in Ncstn KO and WT SSPC

```{r, echo=FALSE}
COI_2 <- readRDS(system.file("extdata", "COI_2.rds", package = "PathwayEmbed"))
```

```{r}
# Extract gene expression matrix and cell meatadata for the input
COI_2_matrix <- as.matrix(GetAssayData(COI_2, assay = "RNA", layer = "data"))
COI_2_metadata <- COI_2@meta.data

# Load Pathway Data
ListPathway("NOTCH")
# Choose "NOTCH_JAG1" as the data comes from mouse cells and "NOTCH_JAG1_24H" from human cells to compare
pathwaydata_mus <- LoadPathway("NOTCH_JAG1", "mouse")
pathwaydata_24hr <- LoadPathway("NOTCH_JAG1_24H", "mouse")

# Datapreprocess -using the matrix already extracted
matrix_mus <- DataPreProcess(COI_2_matrix, pathwaydata_mus, Seurat.object = FALSE)
matrix_24hr <- DataPreProcess(COI_2_matrix, pathwaydata_24hr, Seurat.object = FALSE)

# Pathway Max and Min using the two different Pathway data 
pathwaystat_mus <- PathwayMaxMin(matrix_mus, pathwaydata_mus)
pathwaystat_24hr <- PathwayMaxMin(matrix_24hr, pathwaydata_24hr)

# Calculate Score
Notch_score_mus <- ComputeCellData(matrix_mus, pathwaystat_mus)
Notch_score_24hr <- ComputeCellData(matrix_24hr, pathwaystat_24hr)

# Map metadata back so we can visualize downstream
to_plot_1 <- PreparePlotData(COI_2_metadata, Notch_score_mus, "condition")
to_plot_2 <- PreparePlotData(COI_2_metadata, Notch_score_24hr, "condition")

# Check their comparisons
CalculatePercentage(to_plot_1, "condition")
CalculatePercentage(to_plot_2, "condition")
# Both pathway matrix show that KO has lower activation 

# Define Cohens'd
cohens_d_1 <- round(CalculatePercentage(to_plot_1, "condition")$cohens_d[1], 3)
cohens_d_2 <- round(CalculatePercentage(to_plot_2, "condition")$cohens_d[1], 3) # larger

# Visualization
PlotPathway(to_plot_1, "Notch (Mouse datasets)", "condition", c("#FFDAB9", "#A3BFD9")) + annotate(
    "text",
    x     = Inf, y = Inf,
    label = paste0("Cohen's d = ", cohens_d_1),
    hjust = 1.1, vjust = 1.5,
    size  = 3.5,
    color = "black"
  ) 
PlotPathway(to_plot_2, "Notch (24hr Human cell activation", "condition", c("#FFDAB9", "#A3BFD9")) + annotate(
    "text",
    x     = Inf, y = Inf,
    label = paste0("Cohen's d = ", cohens_d_2),
    hjust = 1.1, vjust = 1.5,
    size  = 3.5,
    color = "black"
  ) 


```

## Data Preprocess For Middle-Age vs Young cells

```{r, eval=FALSE}
# Load Matrix Data Downloaded
wt_data <- Read10X(
  "~/Desktop/Intracellular Pathway/Test data set/WT",
  gene.column = 2,
  cell.column = 1,
  unique.features = TRUE,
  strip.suffix = FALSE
)
WT_object <- CreateSeuratObject(counts = wt_data)

young_data <- Read10X(
  "~/Desktop/Intracellular Pathway/Test data set/Young",
  gene.column = 2,
  cell.column = 1,
  unique.features = TRUE,
  strip.suffix = FALSE
)
Young_object <- CreateSeuratObject(counts = young_data)

# Assign Age Group
WT_object$Age <- 'MiddleAge'
Young_object$Age <- 'Young'

# Filter
# mitochondria percent
WT_object[["percent.mt"]] <- PercentageFeatureSet(WT_object , pattern = "^mt-")
Young_object[["percent.mt"]] <- PercentageFeatureSet(Young_object , pattern = "^mt-")

# Apply filtering: Keep cells with at least 100 genes and <30% mitochondrial content
WT_object_filtered<- subset(WT_object , subset = nFeature_RNA >= 100 & percent.mt < 30)
Young_object_filtered<- subset(Young_object , subset = nFeature_RNA >= 100 & percent.mt < 30)

# Check the number of remaining cells
dim(WT_object_filtered) #[1] 32285  8842
dim(Young_object_filtered) # [1] 32285 15232

# Merge
merge_object <- merge(WT_object_filtered, Young_object_filtered, add.cells.id = c('WT', 'Young'))
merge_object[['RNA']] <- JoinLayers(merge_object[['RNA']])

# Normalization
merge_object <- NormalizeData(merge_object)

## Remove genes detected in fewer than 5 cells
merge_object  <- subset(merge_object , features = rownames(merge_object )[Matrix::rowSums(GetAssayData(merge_object , slot = "counts") > 0) > 5])

# Identiy Highly variable genes
merge_object  <- FindVariableFeatures(merge_object , selection.method = "vst", nfeatures = 3000)

# Scale Data
merge_object <- ScaleData(merge_object)

# PCA
merge_object <- RunPCA(merge_object)
ElbowPlot(merge_object)  # Check how many PCs to use
merge_object <- FindNeighbors(merge_object, dims = 1:10)
merge_object <- FindClusters(merge_object, resolution = 1)
merge_object <- RunUMAP(merge_object, dims = 1:10)

# Visualize
DimPlot(merge_object, reduction = "umap", group.by = "Age", label = TRUE)
DimPlot(merge_object, reduction = "umap", label = TRUE)

# Optional
saveRDS(merge_object, file = "Middle_age_object.rds")

# Load
Middle_age_object <- readRDS("Middle_age_object.rds")

# Identify Skeletal Stem Cells
FeaturePlot(Middle_age_object, reduction = "umap", feature = "Cxcl12", order = TRUE) #11 SSPC Marker in the paper
cluster_of_interest <- subset(Middle_age_object, subset = seurat_clusters == 11) 

```

## Notch Signaling Examination in Middle-Age vs Young cells

```{r, echo=FALSE}
cluster_of_interest <- readRDS(system.file("extdata", "cluster_of_interest.rds", package = "PathwayEmbed"))
```


```{r}
# For pathway data, we called the pathwaydata_24hr that loaded in preivous example
# pathwaydata_24hr <- LoadPathway("NOTCH_JAG1_24H", "mouse")

# Datapreprocess -using Seurat Objct this time 
matrix_notch <- DataPreProcess(cluster_of_interest, pathwaydata_24hr, Seurat.object = TRUE)

# Pathway Max and Min 
pathwaystat_notch <- PathwayMaxMin(matrix_notch, pathwaydata_24hr)

# Calculate Score
Notch_score <- ComputeCellData(matrix_notch, pathwaystat_notch)

# Map metadata back so we can visualize downstream
to_plot <- PreparePlotData(cluster_of_interest, Notch_score, "Age", Seurat.object = TRUE)

# Check their comparisons
age_stats <- CalculatePercentage(to_plot, "Age")
# #   group     percentage_on percentage_off cohens_d
#   <chr>             <dbl>          <dbl>    <dbl>
# 1 MiddleAge          55.8           44.2    0.605
# 2 Young              26.1           73.9    0.605

# Extract stats
cohens_d <- round(age_stats$cohens_d[1], 3)
middleage_on <- age_stats$percentage_on[age_stats$group == "MiddleAge"]
young_on     <- age_stats$percentage_on[age_stats$group == "Young"]

# Build annotation label
annotation_label <- paste0(
  "MiddleAge: ", middleage_on, "% on | ", round(100 - middleage_on, 1), "% off\n",
  "Young: ",     young_on,     "% on | ", round(100 - young_on, 1),     "% off\n",
  "Cohen's d = ", cohens_d
)

# Visualization
PlotPathway(to_plot, "Notch", "Age", c("orange", "#6baed6")) + annotate(
  "text",
  x     = Inf, y = Inf,
  label = annotation_label,
  hjust = 1.1, vjust = 1.5,
  size  = 3.5,
  color = "black"
)
```



```
