---
title: "Spatial Pathway Visualization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatial Pathway Visualization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)

# load library
library(PathwayEmbed)
library(Seurat)
library(ggplot2)
library(viridis)
library(cowplot)

```

## Spatial data load and process

The files can be downloaded from figshare via below link:

Huang, Yaqing (2025). dat3.with.niches.norm.Robj. figshare. Dataset. https://doi.org/10.6084/m9.figshare.29649995.v1

Huang, Yaqing (2025). dat4.with.niches.norm.Robj. figshare. Dataset. https://doi.org/10.6084/m9.figshare.29649989.v1

Huang, Yaqing (2025). dat1.with.niches.norm.Robj. figshare. Dataset. https://doi.org/10.6084/m9.figshare.29649992.v1

Huang, Yaqing (2025). dat2.with.niches.norm.Robj. figshare. Dataset. https://doi.org/10.6084/m9.figshare.29649986.v1

```{r load data, eval=FALSE}
# load data
load("dat1.with.niches.norm.Robj")
load("dat2.with.niches.norm.Robj")
load("dat3.with.niches.norm.Robj")
load("dat4.with.niches.norm.Robj")

# Merge together
merged_spatial <- merge(
  dat1, y = c(dat2, dat3, dat4))

# Set Default Assay to be "RNA"
DefaultAssay(merged_spatial) <- "RNA"
```

## Compute pathway score

Compute score for Wnt, Notch, Hippo, Tgfb, and HIF-1a pathways for the merged subject using 'ComputeCellData' in PathwayEmbed

```{r score compute, eval=FALSE}

# Compute the score for each pathway
Wnt_mds <- ComputeCellData(merged_spatial, "Wnt", "manhattan", batch.size = 1000)
Notch_mds <- ComputeCellData(merged_spatial, "Notch", "manhattan", batch.size = 1000)
Hippo_mds <- ComputeCellData(merged_spatial, "Hippo", "manhattan", batch.size = 1000)
Tgfb_mds <- ComputeCellData(merged_spatial, "Tgfb", "manhattan", batch.size = 1000)
HIF1a_mds <- ComputeCellData(merged_spatial, "HIF-1a", "manhattan", batch.size = 1000)

# Process the mds
Wnt_to.plot <- PreparePlotData(merged_spatial, Wnt_mds, "timepoint")
Notch_to.plot <- PreparePlotData(merged_spatial, Notch_mds, "timepoint")
Hippo_to.plot <- PreparePlotData(merged_spatial, Hippo_mds, "timepoint")
Tgfb_to.plot <- PreparePlotData(merged_spatial, Tgfb_mds, "timepoint")
HIF1a_to.plot <- PreparePlotData(merged_spatial, HIF1a_mds, "timepoint")

# Combine to list
pathway_list <- list(
  Wnt = Wnt_to.plot,
  Notch = Notch_to.plot,
  Hippo = Hippo_to.plot,
  Tgfb = Tgfb_to.plot,
  HIF1a = HIF1a_to.plot
)

```

## Preparation for groups

```{r, echo = FALSE}
# Load to.plot rds
pathway_list <- readRDS(system.file("extdata", "pathway_list.rds", package = "PathwayEmbed"))

```

```{r plot set up}
# Color set-up
magma_colors <- c("#000004FF", "#721F81FF", "#F1605DFF", "#5A90E6")

# Desired timepoint order
ordered_timepoints <- c("E9.5", "E10.5", "E11.5", "E12.5")

# Reorder timepoint levels
for (name in names(pathway_list)) {
  pathway_list[[name]]$timepoint <- factor(pathway_list[[name]]$timepoint, levels = ordered_timepoints)
}

```

## Plot across different timepoints

```{r}

# Loop through each pathway and generate/save the plot
for (i in seq_along(pathway_list)) {
  # Generate the plot
  p <- PlotPathway(pathway_list[[i]], names(pathway_list)[i], "timepoint", magma_colors) +
    facet_wrap(~timepoint, ncol = 1)
  print(p)
}


```

## Merge score with original metadata

```{r, eval = FALSE}
# Step 1: Create named score vectors for each pathway
score_list <- lapply(pathway_list, function(df) {
  s <- df$scale
  names(s) <- rownames(df)
  return(s)
})

# Step 2: Add each pathway score to dat1–dat4
for (i in 1:4) {
  dat <- get(paste0("dat", i))  # get dat1, dat2, ...
  
  for (pathway_name in names(score_list)) {
    score_vec <- score_list[[pathway_name]]
    dat[[paste0(pathway_name, "_score")]] <- score_vec[colnames(dat)]
  }
  
  assign(paste0("dat", i), dat)  # assign back to dat1, dat2, etc.
}


# List of Seurat objects
dat_list <- list(dat1, dat2, dat3, dat4)
names(dat_list) <- paste0("dat", 1:4)

# List of pathways
pathways <- names(pathway_list)  # e.g., "Wnt", "Notch", etc.

```

## Extract the coordinates 

```{r, eval=FALSE}
# Function to extract 
extract_pathway_df <- function(seu, pathway, sample_name = "sample") {
  coords <- as.data.frame(Embeddings(seu[["spatial"]]))
  colnames(coords) <- c("x", "y")
  coords$score <- seu[[paste0(pathway, "_score")]][rownames(coords), 1]
  coords$sample <- sample_name
  return(coords)
}

# Set list to save the coordinates
combined_df_lists <- list()

# For loop for all pathways 
for (pathway in pathways) {
  pathway_df_list <- mapply(
    FUN = extract_pathway_df,
    seu = dat_list,
    sample_name = names(dat_list),
    MoreArgs = list(pathway = pathway),
    SIMPLIFY = FALSE
  )
  
  combined_df <- do.call(rbind, pathway_df_list)
  combined_df_lists[[pathway]] <- combined_df
}

```

## Plot the spatial data

```{r, echo=FALSE}
combined_df_lists <- readRDS(system.file("extdata", "combined_df_lists.rds", package = "PathwayEmbed"))

```

```{r}
limits_list <- list(
  Wnt = c(-3, 4),
  Notch = c(-3, 4),
  Hippo = c(-2, 3),
  Tgfb = c(-2, 3),
  HIF1a = c(-3, 5)
)

for (pathway in names(combined_df_lists)) {
  combined_df <- combined_df_lists[[pathway]]
  selected_limits <- limits_list[[pathway]]
  
  p <- ggplot(combined_df, aes(x = x, y = y, color = score)) +
    geom_point(size = 0.3) +
    scale_color_viridis_c(
      option = "magma",
      name = paste0(pathway, "_score"),
      limits = selected_limits,
      oob = scales::squish
    ) +
    scale_y_reverse() +
    coord_fixed() +
    theme_void() +
    theme(legend.position = "right")
  
  print(p)
}

```
