---
title: "TGF-β Pathway Database Construction"
author: "PathwayEmbed Package"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
    theme: flatly
    highlight: tango
    number_sections: true
vignette: >
  %\VignetteIndexEntry{TGF-β Pathway Database Construction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo     = TRUE,
  message  = FALSE,
  warning  = FALSE,
  collapse = TRUE,
  comment  = "#>"
)
```

---

# Overview

This vignette documents the construction of the **TGF-β signaling pathway database** used by `PathwayEmbed` for downstream pathway activity scoring and embedding. The same construction workflow is applied to all pathway databases in the package (e.g., WNT, NOTCH, YAP); the TGF-β pathway is used here as the worked example.

The database integrates:

1. A manually curated pathway gene list organized by functional category (KEGG hsa04350 )
2. Separately curated mouse ortholog lists, with attention to symbols that differ substantially between species
3. Differential expression results from two independent bulk RNA-seq datasets covering multiple species and contexts
4. Per-gene **activity coefficients** (+1 / −1) derived from the direction of regulation under TGF-β stimulation

The final output is a formatted Excel workbook (`TGFB_Pathway_Database.xlsx`) with three sheets: human Day 1, human Day 20, and mouse.

**Datasets used:**

| GEO Accession | Species | Model | Comparison |
|---|---|---|---|
| GSE110021 | Human | Lung fibroblasts | TGFβ1 vs. no TGFβ at Day 1 and Day 20 |
| GSE246932 | Mouse | T cells | TGFβ vs. none (2h) |

---

# Step 1: Curate the Pathway Gene List

## Human Gene List

TGF-β pathway genes were curated from **KEGG hsa04350** and organized into 20 functional categories spanning canonical SMAD signaling, non-canonical branches (MAPK, PI3K/AKT, RHO), and the extracellular ligand/antagonist landscape.

```{r tgfb-genes-human}
tgfb_genes_human <- list(

  Ligands_TGFB = c("TGFB1", "TGFB2", "TGFB3"),

  Ligands_BMP = c("BMP2", "BMP4", "BMP5", "BMP6", "BMP7", "BMP8A", "BMP8B",
                  "BMP10", "BMP15", "GDF5", "GDF6", "GDF7"),

  # NOTE: INHBC and INHBE have no mouse orthologs (see mouse section)
  Ligands_Activin_GDF = c("INHBA", "INHBB", "INHBC", "INHBE",
                          "GDF1", "GDF2", "GDF3", "GDF8", "GDF9", "GDF10",
                          "GDF11", "GDF15", "AMH", "NODAL"),

  Receptors_TypeI   = c("TGFBR1", "ACVR1", "ACVR1B", "ACVR1C",
                        "BMPR1A", "BMPR1B", "ACVRL1"),

  Receptors_TypeII  = c("TGFBR2", "ACVR2A", "ACVR2B", "BMPR2", "AMHR2"),

  # Type III receptors act as co-receptors/presenters
  Receptors_TypeIII = c("TGFBR3", "ENG"),

  # R-SMADs: phosphorylated by type I receptors; SMAD1/5/9 for BMP arm,
  # SMAD2/3 for TGF-β/Activin arm
  SMAD_Regulated  = c("SMAD1", "SMAD2", "SMAD3", "SMAD5", "SMAD9"),
  SMAD_Common     = c("SMAD4"),
  SMAD_Inhibitory = c("SMAD6", "SMAD7"),

  # SARA (ZFYVE9) anchors SMAD2/3 at the receptor membrane
  SMAD_Anchors = c("SARA", "ZFYVE9"),

  Coactivators = c("CREBBP", "EP300", "SKI", "SKIL", "CITED1", "CITED2",
                   "FOXH1", "FOXO1", "FOXO3", "RUNX1", "RUNX2", "RUNX3",
                   "SP1", "JUN", "FOS", "ATF2"),

  Corepressors = c("TGIF1", "TGIF2", "SKI", "SKIL", "SIN3A", "NCOR1",
                   "NCOR2", "HDAC1", "HDAC2", "HDAC3"),

  E3_Ligases = c("SMURF1", "SMURF2", "NEDD4L", "WWP1", "WWP2",
                 "RNF12", "STUB1", "TRIM33"),

  Transcriptional_Targets = c(
    "SERPINE1", "SMAD7", "CDKN1A", "CDKN2B",
    "MYC", "SNAI1", "SNAI2", "TWIST1", "ZEB1", "ZEB2",
    "VIM", "CDH1", "CDH2", "MMP2", "MMP9",
    "COL1A1", "COL1A2", "COL3A1", "FN1", "ACTA2",
    "CTGF", "TGFB1", "ID1", "ID2", "ID3"
  ),

  Secreted_Antagonists = c(
    "FST", "FSTL1", "FSTL3", "CHRD", "CHRDL1", "CHRDL2",
    "NOG", "GREM1", "GREM2", "NBL1", "BAMBI",
    "LTBP1", "LTBP2", "LTBP3", "LTBP4", "THBS1", "THBS2",
    "DCN", "BGN", "ASPN"
  ),

  MAPK_Pathway = c(
    "MAP3K7", "TAB1", "TAB2", "TAB3",
    "MAP2K3", "MAP2K4", "MAP2K6", "MAP2K7",
    "MAPK1", "MAPK3", "MAPK8", "MAPK9", "MAPK10",
    "MAPK11", "MAPK12", "MAPK13", "MAPK14"
  ),

  PI3K_AKT_Pathway = c("PIK3CA", "PIK3CB", "PIK3CD", "PIK3R1", "PIK3R2",
                       "AKT1", "AKT2", "AKT3", "MTOR", "RHEB"),

  RHO_Pathway = c("RHOA", "RAC1", "CDC42", "ROCK1", "ROCK2",
                  "TGFBR1", "PAK1", "PAK2", "LIMK1", "LIMK2"),

  Latency_Activation = c(
    "LTBP1", "LTBP2", "LTBP3", "LTBP4",
    "LRRC32", "LRRC33", "THBS1", "ITGAV", "ITGB1",
    "ITGB3", "ITGB5", "ITGB6", "ITGB8", "MMP2", "MMP9"
  ),

  Phosphatases = c("PPM1A", "PPP1CA", "PPP1CB", "PPP1CC", "MTMR4"),

  Nuclear_Transport = c("IMPORTIN7", "IMPORTIN8", "XPO1", "RAN"),

  Other_Regulators = c(
    "DAXX", "FNTA", "FKBP1A", "ELF", "YAP1", "TAZ",
    "PMEPA1", "TRIM33", "EVI1", "BCOR", "DRAP1", "MAML1",
    "PPP2CA", "PPP2R1A", "STRAP", "CDKN1B", "CDKN1C",
    "RBL1", "RBL2", "E2F4", "E2F5", "COPS5"
  )
)

# Genes per category
sapply(tgfb_genes_human, length)
cat("Total unique human TGF-β genes:", length(unique(unlist(tgfb_genes_human))), "\n")
```

## Mouse Gene List

> ⚠️ **Mouse symbols are NOT simply lowercased human symbols.** The mouse list was curated manually with special attention to genes that have completely different symbols between species. Key differences are highlighted below.

```{r tgfb-genes-mouse}
# Critical human → mouse symbol differences:
# RNF12       → Rlim          (E3_Ligases)
# IMPORTIN7   → Ipo7          (Nuclear_Transport)
# IMPORTIN8   → Ipo8          (Nuclear_Transport)
# TAZ (WWTR1) → Wwtr1         (Other_Regulators)
# EVI1 (MECOM)→ Mecom         (Other_Regulators)
# SARA        → removed (protein name, not gene; gene is Zfyve9)
# INHBC, INHBE→ removed (no mouse ortholog)
# ELF         → removed (ambiguous; could be Elf1–5)

tgfb_genes_mouse <- list(

  Ligands_TGFB        = c("Tgfb1", "Tgfb2", "Tgfb3"),

  Ligands_BMP         = c("Bmp2", "Bmp4", "Bmp5", "Bmp6", "Bmp7", "Bmp8a", "Bmp8b",
                          "Bmp10", "Bmp15", "Gdf5", "Gdf6", "Gdf7"),

  # INHBC and INHBE have no mouse orthologs and are excluded
  Ligands_Activin_GDF = c("Inhba", "Inhbb",
                          "Gdf1", "Gdf2", "Gdf3", "Gdf8", "Gdf9", "Gdf10",
                          "Gdf11", "Gdf15", "Amh", "Nodal"),

  Receptors_TypeI     = c("Tgfbr1", "Acvr1", "Acvr1b", "Acvr1c",
                          "Bmpr1a", "Bmpr1b", "Acvrl1"),

  Receptors_TypeII    = c("Tgfbr2", "Acvr2a", "Acvr2b", "Bmpr2", "Amhr2"),

  Receptors_TypeIII   = c("Tgfbr3", "Eng"),

  SMAD_Regulated      = c("Smad1", "Smad2", "Smad3", "Smad5", "Smad9"),
  SMAD_Common         = c("Smad4"),
  SMAD_Inhibitory     = c("Smad6", "Smad7"),

  # SARA is a protein name for ZFYVE9; only the gene symbol is used
  SMAD_Anchors        = c("Zfyve9"),

  Coactivators        = c("Crebbp", "Ep300", "Ski", "Skil", "Cited1", "Cited2",
                          "Foxh1", "Foxo1", "Foxo3", "Runx1", "Runx2", "Runx3",
                          "Sp1", "Jun", "Fos", "Atf2"),

  Corepressors        = c("Tgif1", "Tgif2", "Ski", "Skil", "Sin3a", "Ncor1",
                          "Ncor2", "Hdac1", "Hdac2", "Hdac3"),

  # RNF12 (human) = Rlim (mouse) — different symbol!
  E3_Ligases          = c("Smurf1", "Smurf2", "Nedd4l", "Wwp1", "Wwp2",
                          "Rlim", "Stub1", "Trim33"),

  Transcriptional_Targets = c(
    "Serpine1", "Smad7", "Cdkn1a", "Cdkn2b",
    "Myc", "Snai1", "Snai2", "Twist1", "Zeb1", "Zeb2",
    "Vim", "Cdh1", "Cdh2", "Mmp2", "Mmp9",
    "Col1a1", "Col1a2", "Col3a1", "Fn1", "Acta2",
    "Ctgf", "Tgfb1", "Id1", "Id2", "Id3"
  ),

  Secreted_Antagonists = c(
    "Fst", "Fstl1", "Fstl3", "Chrd", "Chrdl1", "Chrdl2",
    "Nog", "Grem1", "Grem2", "Nbl1", "Bambi",
    "Ltbp1", "Ltbp2", "Ltbp3", "Ltbp4", "Thbs1", "Thbs2",
    "Dcn", "Bgn", "Aspn"
  ),

  MAPK_Pathway        = c("Map3k7", "Tab1", "Tab2", "Tab3",
                          "Map2k3", "Map2k4", "Map2k6", "Map2k7",
                          "Mapk1", "Mapk3", "Mapk8", "Mapk9", "Mapk10",
                          "Mapk11", "Mapk12", "Mapk13", "Mapk14"),

  PI3K_AKT_Pathway    = c("Pik3ca", "Pik3cb", "Pik3cd", "Pik3r1", "Pik3r2",
                          "Akt1", "Akt2", "Akt3", "Mtor", "Rheb"),

  RHO_Pathway         = c("Rhoa", "Rac1", "Cdc42", "Rock1", "Rock2",
                          "Tgfbr1", "Pak1", "Pak2", "Limk1", "Limk2"),

  Latency_Activation  = c("Ltbp1", "Ltbp2", "Ltbp3", "Ltbp4",
                          "Lrrc32", "Lrrc33", "Thbs1", "Itgav", "Itgb1",
                          "Itgb3", "Itgb5", "Itgb6", "Itgb8", "Mmp2", "Mmp9"),

  Phosphatases        = c("Ppm1a", "Ppp1ca", "Ppp1cb", "Ppp1cc", "Mtmr4"),

  # IMPORTIN7 = Ipo7, IMPORTIN8 = Ipo8 in mouse
  Nuclear_Transport   = c("Ipo7", "Ipo8", "Xpo1", "Ran"),

  # TAZ = Wwtr1 (mouse), EVI1 = Mecom (mouse)
  Other_Regulators    = c("Daxx", "Fnta", "Fkbp1a",
                          "Yap1", "Wwtr1", "Pmepa1", "Trim33",
                          "Mecom", "Bcor", "Drap1", "Maml1",
                          "Ppp2ca", "Ppp2r1a", "Strap", "Cdkn1b", "Cdkn1c",
                          "Rbl1", "Rbl2", "E2f4", "E2f5", "Cops5")
)

cat("Total unique mouse TGF-β genes:", length(unique(unlist(tgfb_genes_mouse))), "\n")
```

## Flatten to Data Frames

Both lists are converted to flat two-column data frames (`gene`, `category`) for downstream joining operations.

```{r flatten}
flatten_gene_list <- function(gene_list) {
  do.call(rbind, lapply(names(gene_list), function(cat) {
    data.frame(gene = gene_list[[cat]], category = cat, stringsAsFactors = FALSE)
  }))
}

tgfb_genes_human_flat <- flatten_gene_list(tgfb_genes_human)
tgfb_genes_mouse_flat <- flatten_gene_list(tgfb_genes_mouse)

head(tgfb_genes_human_flat)
```

---

# Step 2: Bulk RNA-seq Differential Expression

## Human — Lung Fibroblasts (GSE110021)

Human WI-38 fibroblasts were treated with TGFβ1 and profiled at **Day 1** (acute response) and **Day 20** (chronic/fibrotic response). The dataset provides voom-normalized log-counts, so limma is applied directly without further normalization.

```{r human-rnaseq, eval=FALSE}
library(limma)

# Load voom-normalized expression matrix
data_fibroblasts <- read.table(
  "GSE110021_counts.voom.annot.txt.gz",
  header = TRUE, sep = "\t", row.names = 1
)

# Separate expression values from annotation columns
expr <- as.matrix(data_fibroblasts[, -(1:2)])
rownames(expr) <- data_fibroblasts$GeneSymbol

# Sample metadata: 24 samples total, 12 per treatment, 6 per timepoint per treatment
sample_info <- data.frame(
  sample    = colnames(expr),
  treatment = rep(c("noTGFb", "TGFb"), each = 12),
  timepoint = rep(c("D1", "D20"), each = 6, times = 2)
)

# Design matrix with all four groups
design <- model.matrix(~ 0 + treatment:timepoint, data = sample_info)
colnames(design) <- c("D1_noTGFb", "D1_TGFb", "D20_noTGFb", "D20_TGFb")

fit <- lmFit(expr, design)

# Contrasts: TGFb vs. control at each timepoint
contrast_matrix <- makeContrasts(
  D1_TGFb_vs_noTGFb  = D1_TGFb  - D1_noTGFb,
  D20_TGFb_vs_noTGFb = D20_TGFb - D20_noTGFb,
  levels = design
)

fit2 <- contrasts.fit(fit, contrast_matrix)
fit2 <- eBayes(fit2)

# Extract full results tables
res_D1  <- topTable(fit2, coef = "D1_TGFb_vs_noTGFb",  number = Inf, sort.by = "p")
res_D20 <- topTable(fit2, coef = "D20_TGFb_vs_noTGFb", number = Inf, sort.by = "p")

# Add gene descriptions
res_D1$Description  <- data_fibroblasts$Description[match(rownames(res_D1),  data_fibroblasts$GeneSymbol)]
res_D20$Description <- data_fibroblasts$Description[match(rownames(res_D20), data_fibroblasts$GeneSymbol)]

write.csv(res_D1,  "DEG_D1_TGFb_vs_noTGFb.csv",  row.names = TRUE)
write.csv(res_D20, "DEG_D20_TGFb_vs_noTGFb.csv", row.names = TRUE)
```

## Mouse — T Cells (GSE246932)

Mouse CD8+ T cells were treated with TGFβ for 2 hours. Raw count data required DESeq2 for normalization and differential testing.

```{r mouse-rnaseq, eval=FALSE}
library(DESeq2)
library(dplyr)

# Load raw counts
counts_data <- read.csv(
  "GSE246932_220809-P490-1_RawGeneCounts.csv.gz",
  header = TRUE, check.names = FALSE
)

# Remove SIINFEKL peptide-treated samples (not relevant to TGFb comparison)
counts_data <- counts_data[, !grepl("SIINFEKL", colnames(counts_data))]

# Separate gene annotation from count columns
gene_anno  <- counts_data[, 1:4]
counts_only <- counts_data[, -(1:4)]

# Collapse duplicate gene IDs by summing
counts_collapsed <- counts_only %>%
  mutate(geneId = gene_anno$geneId) %>%
  group_by(geneId) %>%
  summarise(across(where(is.numeric), sum), .groups = "drop")

counts_mat <- as.matrix(counts_collapsed[, -1])
rownames(counts_mat) <- counts_collapsed$geneId

# Build DESeq2 object
sample_names <- colnames(counts_mat)
coldata <- data.frame(
  treatment = factor(
    ifelse(grepl("TGFb", sample_names), "TGFb", "none"),
    levels = c("none", "TGFb")
  ),
  row.names = sample_names
)

dds <- DESeqDataSetFromMatrix(
  countData = counts_mat,
  colData   = coldata,
  design    = ~ treatment
)
dds <- dds[rowSums(counts(dds)) > 10, ]  # low-count filter
dds <- DESeq(dds)

res    <- results(dds, contrast = c("treatment", "TGFb", "none"))
res_df <- as.data.frame(res)

# Map gene names back from annotation
res_df$geneId      <- rownames(res_df)
res_df$geneName    <- gene_anno$geneName[match(res_df$geneId, gene_anno$geneId)]
res_df$description <- gene_anno$description[match(res_df$geneId, gene_anno$geneId)]

library(openxlsx)
write.xlsx(res_df, file = "Mus_TGFb_vs_none.xlsx", rowNames = FALSE)
```

---

# Step 3: Filter for TGF-β Pathway Genes

DEG results are joined against the curated gene lists, retaining only TGF-β pathway members that reach statistical significance (adjusted *p* < 0.05).

```{r filter, eval=FALSE}
library(readr)
library(readxl)
library(dplyr)

human_d1  <- read_csv("DEG_D1_TGFb_vs_noTGFb.csv")
human_d20 <- read_csv("DEG_D20_TGFb_vs_noTGFb.csv")
mouse_df  <- read_excel("Mus_TGFb_vs_none.xlsx")

# Human Day 1
human_d1_results <- human_d1 %>%
  filter(ID %in% tgfb_genes_human_flat$gene) %>%
  left_join(tgfb_genes_human_flat, by = c("ID" = "gene")) %>%
  filter(!is.na(adj.P.Val) & adj.P.Val < 0.05) %>%
  arrange(adj.P.Val)

# Human Day 20
human_d20_results <- human_d20 %>%
  filter(ID %in% tgfb_genes_human_flat$gene) %>%
  left_join(tgfb_genes_human_flat, by = c("ID" = "gene")) %>%
  filter(!is.na(adj.P.Val) & adj.P.Val < 0.05) %>%
  arrange(adj.P.Val)

# Mouse
mouse_df_results <- mouse_df %>%
  filter(geneName %in% tgfb_genes_mouse_flat$gene) %>%
  left_join(tgfb_genes_mouse_flat, by = c("geneName" = "gene")) %>%
  filter(!is.na(padj) & padj < 0.05) %>%
  arrange(padj)

cat("Human D1  pathway genes (padj<0.05):", nrow(human_d1_results),  "\n")
cat("Human D20 pathway genes (padj<0.05):", nrow(human_d20_results), "\n")
cat("Mouse     pathway genes (padj<0.05):", nrow(mouse_df_results),  "\n")
```

---

# Step 4: Assign Activity Coefficients

## Rationale

Each gene receives a **direction coefficient** reflecting its behavior under TGFβ activation:

| Coefficient | Meaning |
|:-----------:|---------|
| **+1** | Upregulated with TGFβ — consistent with pathway activation |
| **−1** | Downregulated with TGFβ — consistent with pathway inhibition |

These coefficients are used by `PathwayEmbed` to compute a **signed pathway activity score**: the weighted sum of a sample's expression values multiplied by their coefficients, giving a single number that reflects the net direction and magnitude of TGF-β activity.

The function also supports inhibitor experiments via `TGFB_activation = FALSE`, which flips the sign logic — a gene downregulated by an inhibitor is inferred to be positively regulated under activation.

```{r coef-function}
assign_coefficient_tgfb <- function(log2FoldChange,
                                    fc_threshold    = 0,
                                    TGFB_activation = TRUE) {
  if (is.na(log2FoldChange)) return(0)

  if (TGFB_activation) {
    if      (log2FoldChange >  fc_threshold) return( 1)
    else if (log2FoldChange < -fc_threshold) return(-1)
    else                                      return( 0)
  } else {
    # Inhibitor data: flip direction
    if      (log2FoldChange < -fc_threshold) return( 1)
    else if (log2FoldChange >  fc_threshold) return(-1)
    else                                      return( 0)
  }
}
```

> **Threshold note:** `fc_threshold = 0` assigns a coefficient to any non-zero fold-change. For stricter databases, use `fc_threshold = 0.5` to restrict to genes with |log₂FC| ≥ 0.5, focusing on more robust responses.

```{r apply-coef, eval=FALSE}
human_d1_results <- human_d1_results %>%
  mutate(coef = sapply(logFC, assign_coefficient_tgfb, fc_threshold = 0))

human_d20_results <- human_d20_results %>%
  mutate(coef = sapply(logFC, assign_coefficient_tgfb, fc_threshold = 0))

mouse_df_results <- mouse_df_results %>%
  mutate(coef = sapply(log2FoldChange, assign_coefficient_tgfb, fc_threshold = 0))
```

---

# Step 5: Export to Excel

The final database is exported as a formatted `.xlsx` workbook with a styled header row, frozen pane, and auto-sized columns.

```{r export, eval=FALSE}
library(openxlsx)

# --- Helper: prepare one sheet's data frame ---
prepare_sheet <- function(results, gene_col, logfc_col) {
  results %>%
    dplyr::select(
      Gene_Symbol    = !!sym(gene_col),
      Category       = category,
      Coefficient    = coef,
      log2FoldChange = !!sym(logfc_col)
    ) %>%
    as.data.frame()
}

human_d1_sheet  <- prepare_sheet(human_d1_results,  "ID",       "logFC")
human_d20_sheet <- prepare_sheet(human_d20_results, "ID",       "logFC")
mouse_df_sheet  <- prepare_sheet(mouse_df_results,  "geneName", "log2FoldChange")

# --- Helper: write one formatted sheet ---
add_pathway_sheet <- function(wb, sheet_name, df) {

  addWorksheet(wb, sheet_name)

  writeData(wb, sheet_name, df,
    headerStyle = createStyle(
      fontSize = 11, fontColour = "#FFFFFF", halign = "center",
      fgFill = "#4472C4", border = "TopBottom",
      fontName = "Arial", textDecoration = "bold"
    )
  )

  freezePane(wb, sheet_name, firstRow = TRUE)
  tryCatch(
    setColWidths(wb, sheet_name, cols = seq_len(ncol(df)), widths = "auto"),
    error = function(e)
      setColWidths(wb, sheet_name, cols = seq_len(ncol(df)), widths = 15)
  )
}

# --- Build workbook ---
wb <- createWorkbook()
add_pathway_sheet(wb, "TGFB_Human_D1",  human_d1_sheet)
add_pathway_sheet(wb, "TGFB_Human_D20", human_d20_sheet)
add_pathway_sheet(wb, "TGFB_Mouse",     mouse_df_sheet)

saveWorkbook(wb, "TGFB_Pathway_Database.xlsx", overwrite = TRUE)
cat("Saved: TGFB_Pathway_Database.xlsx\n")
```

---

# Final Database Structure

Each sheet in `TGFB_Pathway_Database.xlsx` contains:

| Column | Description |
|--------|-------------|
| `Gene_Symbol` | HGNC (human) or MGI (mouse) gene symbol |
| `Category` | Functional category from KEGG + curation |
| `Coefficient` | +1 (activated) or −1 (repressed) under TGFβ; filtered by padj < 0.05 |
| `log2FoldChange` | Log₂ fold-change from the corresponding DEG analysis |

Sheets produced:

| Sheet | Species | Context | Method |
|-------|---------|---------|--------|
| `TGFB_Human_D1` | Human | Lung fibroblasts, Day 1 | limma / voom |
| `TGFB_Human_D20` | Human | Lung fibroblasts, Day 20 | limma / voom |
| `TGFB_Mouse` | Mouse | CD8+ T cells, 2h | DESeq2 |

---

# Important Curation Notes

The following caveats should be kept in mind when extending or modifying the database:

**Genes excluded from the mouse list due to no ortholog:**
`INHBC`, `INHBE` — these activin subunits are primate-specific and have no mouse ortholog.

**Gene name ambiguities resolved:**
`SARA` is a protein alias for the gene `ZFYVE9` and should not be used as a gene symbol. `ELF` is ambiguous (could refer to `ELF1` through `ELF5`) and was removed from the mouse list.

**Genes with completely different human/mouse symbols:**

| Human | Mouse | Category |
|-------|-------|----------|
| RNF12 | Rlim | E3_Ligases |
| IMPORTIN7 | Ipo7 | Nuclear_Transport |
| IMPORTIN8 | Ipo8 | Nuclear_Transport |
| TAZ (WWTR1) | Wwtr1 | Other_Regulators |
| EVI1 (MECOM) | Mecom | Other_Regulators |

Always verify mouse gene symbols against [MGI](http://www.informatics.jax.org/) before adding new genes.

---

# Session Information

```{r session-info}
sessionInfo()
```
